The brightness and spatial distributions of terrestrial radio sources

### 2013-01-09 articles/Article-RFI-distribution/article-rfi-distribution.tex
Some comments of Leon

\begin{abstract}
Faint undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are stationary and coherently present over time. Thereby, they might obstruct the detection of the weak astronomical signals of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest objects in the Universe. We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR. The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. These observations almost reach the thermal noise and are thought not to be limited by RFI. Therefore, we conclude that it is likely that most RFI sources do not add coherently, in which case RFI will not prevent LOFAR from detecting the EoR.
\end{abstract}

@@

Typical radio observations record a few hours of data, and the results are integrated. In these cases, it often suffices to only excise interference that is apparent and thus above the noise, to reach the thermal noise limit of the instrument. A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, weak interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be coherently present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for the 21-cm Epoch of Reionisation (EoR) experiments, because they involve long integration times. Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}. For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) \citep{de-bruyn-eor-ursi-2011}.

@@

In every dynamic spectrum we can measure the number of times that the flux density is within a particular range. Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity. We will refer to this quantity with the statistical term ``frequency density'' (not to be confused with the physical radiation frequency). We will now start by deriving a prediction of the frequency density function of ground-based interfering sources. Consider an interfering point source of strength $I$. This source is observed by an interferometer that consists of two antennas or stations with gains $g_1,g_2$, which include all instrumental effects. The antennas are located at distances $r_1,r_2$ from the source. The interferometer will record an apparent instantaneous strength $\mathcal{S}$ of
\begin{equation}
 \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{r_1 r_2},
\end{equation}
with $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$. Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth. The formula represents a spherically propagating wave in free space. We will limit our analysis to cross-correlated antennas; the auto-correlations will be ignored.

@@

Now, we can treat the interferometer geometrically as a single point, as both antennas will see the same distribution. Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and interferometric gain $g$ as
\begin{equation} \label{eq:amplitude-fall-off-in-free-space}
 S(r) = \frac{Ig}{r^2}.
\end{equation}

@@

The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can now be calculated with
\begin{align} \label{eq:cumulative-rfi-distribution}
 F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{S}^{-1}(r))\\
\notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{S})\\
\notag = & N-\frac{\rho \pi Ig}{S}
\end{align}
where $\mathcal{S}^{-1}(r)$ is the inverse of $\mathcal{S}(r)$, i.e., the function that returns the distance $r$ for a given amplitude $S$. Finally, the differential frequency density can be calculated by taking the derivative,
\begin{align} \label{eq:two-dimensional-distribution}
 f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{\rho \pi Ig}{S^2}.
\end{align}

Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we predict to see a power law with a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane.

\subsection{Spherical case}
If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes. In this case, the RFI sources are distributed over the surface of a sphere. If we assume their difference in height is negligible, and by using the formula for the surface area of a hemisphere and some basic geometry, one finds that
\begin{equation}\label{eq:cumulative-spherical}
\tilde{F}_{\textrm{distance} \le r} = \rho 2\pi R^2\left(1-\cos(2\arcsin \frac{r}{2R})\right) \hspace{0.4cm}\textrm{with } r \le 2R,
\end{equation}
with $R$ the radius of the Earth. Eqs.~\eqref{eq:cumulative-distance-function} and \eqref{eq:cumulative-spherical} have the same value over the range they can be evaluated. Therefore, the spherical and 2-dimensional cases lead to a distribution with equal power law. Evaluating Eq.~\eqref{eq:cumulative-spherical} is limited to $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth. This causes distant sources to settle below the horizon. Nevertheless, some of those sources might still be visible due to reflection of the ionosphere and other (semi-)spherical propagation effects.

\subsection{Propagation effects}
So far, we have assumed that the electromagnetic radiation propagates through free space, resulting in an $r^{-2}$ fall-off. In reality, the radiation will be affected by complicated propagational effects. Because of the irregular surface of the Earth (including urban areas) and the absence of a direct line of sight between transmitter and receiver, the propagation mode might be indirect. Multiple indirect paths might be formed by reflection, refraction or diffraction of the electromagnetic wave.

@@

\subsection{Noise contribution} \label{sec:histogram-noise}
The full measured distribution will consist of the power-law distribution combined with that of the noise and the celestial signal. For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong sources. Noise, however, will have a contribution. The real and imaginary components of receiver noise are independent and identically Gaussian distributed. Consequently, an amplitude $x$ will be Rayleigh distributed: 
\begin{equation}\label{eq:rayleigh-formula}
f_\textrm{noise}(x) =
\begin{cases}
\frac{x}{\sigma^2} e^{\frac{-x^2}{2\sigma^2}} & x > 0, \\
0 & \textrm{otherwise.}
\end{cases}
\end{equation}
Because most of the samples will be unaffected by RFI, this will be the dominating distribution. The Rayleigh distribution is plotted together with the -2 power-law distribution of Eq.~\eqref{eq:two-dimensional-distribution} in Fig.~\ref{fig:rayleigh-and-rfi-distributions}.

@@

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial. The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by normalizing and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, to draw a sample by using the inversion method, this would require a uniformly distributed variable $\sim U(0,\infty)$. By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-\rho \pi r_L^2$ for $S < Ig/r_L^2$. Consequently, by sampling $x_u$ from a uniform distribution with $0 \le x_u \le N-\rho \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ a power-law distribution.

@@

\section{Data description} \label{sec:dist-data}
We have analysed the distributions of two data sets. Both data sets are 24 h LOFAR RFI surveys and are extensively analysed in
\citet{lofar-radio-environment}.
In one set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz. More stations were used in the LBA set. The specifications of the two sets are listed in Table~\ref{table:dist-data-specs}. The stations that have been used are geometrically spread over an area of about 80~km and 30~km in diameter at maximum for the LBA and HBA sets respectively. For EoR detection experiments, the HBA are more important than the LBA, because they cover the frequency range of the redshifted EoR signal.

@@

>>>>

\begin{abstract}
Faint undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are stationary and coherently present over time. Thereby, they might obstruct the detection of the weak astronomical signals of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest structures in the Universe. We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of visibility data observed with LOFAR, similar to radio source count analyses that are used in cosmology. The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. These observations almost reach the thermal noise and are thought not to be limited by RFI. Therefore, we conclude that it is likely that most RFI sources do not add coherently, in which case RFI will not prevent an EoR detection with LOFAR.
\end{abstract}

@@

Typical radio observations record a few hours of data, and the results are integrated. In these cases, it often suffices to only excise interference that is apparent and thus above the noise, to reach the thermal noise limit of the instrument. A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, weak interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be coherently present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for the 21-cm Epoch of Reionisation (EoR) experiments, because they involve long integration times. Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}. For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, van Haarlem et al. 2013, \textit{submitted}).
% TODO correct reference if van Haarlem et al is published

@@

In every dynamic spectrum we can measure the number of times that the flux density is within a particular range. Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity. We will refer to this quantity with the statistical term ``frequency density'' (not to be confused with the physical radiation frequency). We will now start by deriving a prediction of the frequency density function of ground-based interfering sources. Consider an interfering point source of strength $I$ measured in flux density. This source is observed by an interferometer that consists of two antennas or stations with gains $g_1,g_2$, which include all instrumental effects. The antennas are located at distances $r_1,r_2$ from the source. The interferometer will record an apparent instantaneous strength $\mathcal{S}$ of
\begin{equation}
 \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{r_1 r_2},
\end{equation}
with (amplitude) gains $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$. Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth. The formula represents a spherically propagating wave in free space. We will limit our analysis to cross-correlated antennas; the auto-correlations will be ignored.

@@

Although two antennas do not necessarily observe the same RFI sources, for source count analysis we can treat the interferometer geometrically as a single point, as both antennas will see the same distribution. Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and interferometric gain $g = g_1 g_2$ as
\begin{equation} \label{eq:amplitude-fall-off-in-free-space}
 S(r) = \frac{Ig}{r^2}.
\end{equation}

@@

The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can now be calculated with
\begin{align} \label{eq:cumulative-rfi-distribution}
 F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{R}(S))\\
\notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{S})\\
\notag = & N-\frac{\rho \pi Ig}{S}
\end{align}
where $\mathcal{R}(S)=\mathcal{S}^{-1}$, the inverse of $\mathcal{S}$, i.e., the function that returns the distance $r$ for a given amplitude $S$. Finally, the differential frequency density can be calculated by taking the derivative,
\begin{align} \label{eq:two-dimensional-distribution}
 f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{\rho \pi Ig}{S^2}.
\end{align}

Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we predict to see a power law with a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane. For this, we have assumed that the variables $I$, $g$ and $\rho$ are constants. In reality, these variables have a stochastic nature. The effect of this will be discussed in \S\ref{sec:parameter-variability}.

\subsection{Spherical case}
If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes. In this case, the RFI sources are distributed over the surface of a sphere. If we assume their difference in height is negligible, and by using the formula for the surface area of a hemisphere and some basic geometry, one finds that
\begin{equation}\label{eq:cumulative-spherical}
\tilde{F}_{\textrm{distance} \le r} = \rho 2\pi R^2\left(1-\cos(2\arcsin \frac{r}{2R})\right) \hspace{0.4cm}\textrm{with } r \le 2R,
\end{equation}
with $R$ the radius of the Earth and $r$ the Euclidean distance between the source and the antenna. Eqs.~\eqref{eq:cumulative-distance-function} and \eqref{eq:cumulative-spherical} have the same value over the range they can be evaluated. Therefore, the spherical and 2-dimensional cases lead to a distribution with equal power law. Evaluating Eq.~\eqref{eq:cumulative-spherical} is limited to $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth. This causes distant sources to settle below the horizon. Nevertheless, some of those sources might still be visible due to reflection of the ionosphere and other (semi-)spherical propagation effects.

\subsection{Propagation effects}
So far, we have assumed that the electromagnetic radiation propagates through free space, resulting in an $r^{-2}$ fall-off. In reality, the radiation will be affected by complicated propagational effects. Because of the irregular surface of the Earth (including urban areas) and the absence of a direct line of sight between transmitter and receiver, the propagation mode might be indirect. Multiple indirect paths might be formed that include reflection, refraction, diffraction and absorption of the electromagnetic wave.

@@

\subsection{Noise contribution} \label{sec:histogram-noise}
The full measured distribution will consist of the power-law distribution combined with that of the noise and the celestial signal. For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong sources. For example, the strongest apparent source in the NCP EoR field is around 5 Jy (Yattawata et al. 2013, \textit{submitted}). % TODO fill in details.
The standard deviation of the noise, however, is around 100~Jy on highest LOFAR resolutions, and will have a larger contribution.

The real and imaginary components of receiver noise are independent and identically Gaussian distributed with zero mean and equal variance. Consequently, an amplitude $x$ will be Rayleigh distributed \citep[\S6-2]{papoulis-stochastic-processes}: 
\begin{equation}\label{eq:rayleigh-formula}
f_\textrm{noise}(x) =
\begin{cases}
\frac{x}{\sigma^2} e^{\frac{-x^2}{2\sigma^2}} & x > 0, \\
0 & \textrm{otherwise.}
\end{cases}
\end{equation}
Because most of the samples will be unaffected by RFI, this will be the dominating distribution. The Rayleigh distribution is plotted together with the -2 power-law distribution of Eq.~\eqref{eq:two-dimensional-distribution} in Fig.~\ref{fig:rayleigh-and-rfi-distributions}.

@@

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial. The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by normalizing and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, to draw a sample by using the inversion method, this would require a uniformly distributed variable $\sim U(0,\infty)$. By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-\rho \pi r_L^2$ for $S < Ig/r_L^2$. Consequently, if $x_u$ is sampled from a uniform distribution with $0 \le x_u \le N-\rho \pi r_L^2$, and $x_u$ is the input to the inverse of the cumulative source count function in Eq.~\eqref{eq:cumulative-rfi-distribution}, the result is an amplitude sample $S$ with $S \ge Ig/r_L^2$, and $S\sim$ a power-law distribution.

@@

\section{Data description} \label{sec:dist-data}
We have analysed the distributions of two data sets. Both data sets are 24-h LOFAR RFI surveys and are extensively analysed in
\citet{lofar-radio-environment}. We refer to van Haarlem et al. (2013, \textit{submitted}) for a full description of the capabilities of LOFAR.
% TODO correct reference if van Haarlem et al is published
In one set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz. More stations were used in the LBA set. The specifications of the two sets are listed in Table~\ref{table:dist-data-specs}. The stations that have been used are geometrically spread over an area of about 80~km and 30~km in diameter at maximum for the LBA and HBA sets respectively. For EoR detection experiments, the HBA are more important than the LBA, because they cover the frequency range of the redshifted EoR signal.

@@

% TODO appendix with calculation of 2D case == 3D case?
